Supplement 1

### 2023-06-23 paper/GEB Submission/GEB_title_page.Rmd
Save out tables to create doc by hand.

---
title: Prevailing scenarios of functional change in anthropocene bird and mammal communities
authors:
  - name: "Kari E. A. Norman"
    affiliation: Université de Montréal
    location: Montréal, Quebec
    email: kari.norman@berkeley.edu
  - name: "Perry de Valpine"
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: pdevalpine@berkeley.edu
  - name: "Carl Boettiger"
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: cboettig@berkeley.edu
abstract: |
  Aim: Despite unprecedented environmental change due to anthropogenic pressure, recent work has found increasing species turnover but no overall trend in species diversity through time at the local scale. Functional diversity provides a potentially powerful alternative approach for understanding community composition by linking shifts in species identity to mechanisms of ecosystem processes. Here we present the first multi-taxa, multi-system analysis of functional change through time.

@@

  Main Conclusions: General trends indicate that on the aggregate one type of functional shift is not more prevelant than the other across many taxa, biomes, and realms. At the study-level, there were also a substantial number of time series exhibiting no species or functional change, however the majority of studies showed a shift in a species or functional metric. With no one prevailing scenario of change, it will be critical to link change scenarios to drivers of change, particularly to identify communities with capacity to resist drivers from those not experiencing substantial pressure from a driver.
journal: "Ecology Letters"
date: "`r Sys.Date()`"
bibliography: refs.bib
csl: pnas.csl
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: 
  rticles::arxiv_article:
    keep_tex: true
#  word_document: default
editor_options: 
  markdown: 
    wrap: sentence
---

>>>>

---
title: Prevailing scenarios of functional change in Anthropocene bird and mammal communities
authors:
  - name: "Kari E. A. Norman"
    affiliation: Université de Montréal
    location: Montréal, Quebec
    email: kari.norman@berkeley.edu
  - name: "Perry de Valpine"
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: pdevalpine@berkeley.edu
  - name: "Carl Boettiger"
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: cboettig@berkeley.edu
abstract: |
  Aim: Despite unprecedented environmental change due to anthropogenic pressure, recent work has found increasing species turnover but no overall trend in species diversity through time at the local scale. Functional diversity provides a potentially powerful alternative approach for understanding community composition by linking shifts in species identity to mechanisms of ecosystem processes. Here we present the first multi-taxa, multi-system analysis of functional change through time.

@@

  Main Conclusions: General trends indicate that on the aggregate one type of functional shift is not more prevalant than the other across many taxa, biomes, and realms. At the study-level, there were also a substantial number of time series exhibiting no species or functional change, however the majority of studies showed a shift in a species or functional metric. With no one prevailing scenario of change, it will be critical to link change scenarios to drivers of change, particularly to identify communities with capacity to resist drivers from those not experiencing substantial pressure from a driver.
journal: "Ecology Letters"
date: "`r Sys.Date()`"
bibliography: refs.bib
csl: pnas.csl
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: 
  rticles::arxiv_article:
    keep_tex: true
#  word_document: default
editor_options: 
  markdown: 
    wrap: sentence
---

### 2023-06-23 paper/GEB Submission/GEB_title_page.tex
Save out tables to create doc by hand.

\title{Prevailing scenarios of functional change in anthropocene bird
and mammal communities}

@@

Main Conclusions: General trends indicate that on the aggregate one type
of functional shift is not more prevelant than the other across many
taxa, biomes, and realms. At the study-level, there were also a
substantial number of time series exhibiting no species or functional
change, however the majority of studies showed a shift in a species or
functional metric. With no one prevailing scenario of change, it will be
critical to link change scenarios to drivers of change, particularly to
identify communities with capacity to resist drivers from those not
experiencing substantial pressure from a driver.
\end{abstract}

@@

>>>>

\title{Prevailing scenarios of functional change in Anthropocene bird
and mammal communities}

@@

Main Conclusions: General trends indicate that on the aggregate one type
of functional shift is not more prevalant than the other across many
taxa, biomes, and realms. At the study-level, there were also a
substantial number of time series exhibiting no species or functional
change, however the majority of studies showed a shift in a species or
functional metric. With no one prevailing scenario of change, it will be
critical to link change scenarios to drivers of change, particularly to
identify communities with capacity to resist drivers from those not
experiencing substantial pressure from a driver.
\end{abstract}

@@

\hypertarget{biosketch}{%
\section{Biosketch}\label{biosketch}}

Kari Norman's research is generally focused on understanding the impacts
of anthropogenic change on biodiversity. She brings together existing
data sources to understand broadscale change and develops software tools
to facilitate robust synthesis workflows.

### 2023-06-23 paper/GEB Submission/manuscript_GEB.Rmd
Save out tables to create doc by hand.

---
title: Prevailing scenarios of functional change in anthropocene bird and mammal communities
abstract: |
  Aim: Despite unprecedented environmental change due to anthropogenic pressure, recent work has found increasing species turnover but no overall trend in species diversity through time at the local scale. Functional diversity provides a potentially powerful alternative approach for understanding community composition by linking shifts in species identity to mechanisms of ecosystem processes. Here we present the first multi-taxa, multi-system analysis of functional change through time.

@@

  Main Conclusions: General trends indicate that on the aggregate one type of functional shift is not more prevelant than the other across many taxa, biomes, and realms. At the study-level, there were also a substantial number of time series exhibiting no species or functional change, however the majority of studies showed a shift in a species or functional metric. With no one prevailing scenario of change, it will be critical to link change scenarios to drivers of change, particularly to identify communities with capacity to resist drivers from those not experiencing substantial pressure from a driver.

@@

To test for trends within and between different levels of taxa, biome, and realm we fit separate models with each of those factors added as a predictor to the original model structure.
We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.8.2, @lenth2022].
We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
We also assessed the impact of potential study-level outliers on general trends.
Details on that process can be found in the supplement.
All models in our analysis were fit using the *lme4* (v1.1-30) package in R (v4.2.3) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].

>>>>

---
title: Prevailing scenarios of functional change in Anthropocene bird and mammal communities
abstract: |
  Aim: Despite unprecedented environmental change due to anthropogenic pressure, recent work has found increasing species turnover but no overall trend in species diversity through time at the local scale. Functional diversity provides a potentially powerful alternative approach for understanding community composition by linking shifts in species identity to mechanisms of ecosystem processes. Here we present the first multi-taxa, multi-system analysis of functional change through time.

@@

  Main Conclusions: General trends indicate that on the aggregate one type of functional shift is not more prevalant than the other across many taxa, biomes, and realms. At the study-level, there were also a substantial number of time series exhibiting no species or functional change, however the majority of studies showed a shift in a species or functional metric. With no one prevailing scenario of change, it will be critical to link change scenarios to drivers of change, particularly to identify communities with capacity to resist drivers from those not experiencing substantial pressure from a driver.

@@

To test for trends within and between different levels of taxa, biome, and realm we fit separate models with each of those factors added as a predictor to the original model structure.
We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.8.2, @lenth2022].
We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
All models in our analysis were fit using the *lme4* (v1.1-30) package in R (v4.2.3) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].

### 2023-06-23 paper/supp1_sensitivity_outliers.Rmd
Save out tables to create doc by hand.

---
title: "Sensitivity Analysis"
output:
  pdf_document: default
  html_document: default
---

@@

tibble(sensitivity_test = c(rep("duration", 3), rep("trait coverage", 2), "amphibians"), 
       model_id = c("min3", "min4", "min5", "covg85", "covg90", "with_amph"),
       description = c("Exclude time series with fewer than 3 years",
                       "Exclude time series with fewer than 4 years",
                       "Exclude time series with fewer than 5 years",
                       "Exclude time series with less than 85% of species with trait data",
                       "Exclude time series with less than 90% of species with trait data",
                       "Include Amphibian data"),
       sig_metrics = c("CWM seed diet*, CWM canopy foraging*, CWM aerial foraging, Jaccard", 
       "CWM seed diet*, CWM canopy foraging*, FRic*, Jaccard", 
       "CWM seed diet*, CWM canopy foraging*, Jaccard", rep("Jaccard", 3))
) %>% kable(format = "latex") %>%
  column_spec(3:4, width = "3cm") %>%
  collapse_rows(columns = 1, valign = "middle") %>%
  kable_paper() %>%
  kable_styling(latex_options= c("hold_position"))
```

>>>>

---
title: "Supplement 1"
output:
  #pdf_document: default
  word_document: default
---

# Sensitivity Analysis

@@

tibble(sensitivity_test = c(rep("duration", 3), rep("trait coverage", 2), "amphibians"), 
       model_id = c("min3", "min4", "min5", "covg85", "covg90", "with_amph"),
       description = c("Exclude time series with fewer than 3 years",
                       "Exclude time series with fewer than 4 years",
                       "Exclude time series with fewer than 5 years",
                       "Exclude time series with less than 85% of species with trait data",
                       "Exclude time series with less than 90% of species with trait data",
                       "Include Amphibian data"),
       sig_metrics = c("CWM seed diet*, CWM canopy foraging*, CWM aerial foraging, Jaccard", 
       "CWM seed diet*, CWM canopy foraging*, FRic*, Jaccard", 
       "CWM seed diet*, CWM canopy foraging*, Jaccard", rep("Jaccard", 3))
) %>% 
  #kable(format = "latex", caption = "Summary of the results for the sensitivity analyses, including those for duration of time series, trait coverage, and amphibian inclusion.") %>%
  kable(format = "latex") %>%
  column_spec(3:4, width = "3cm") %>%
  collapse_rows(columns = 1, valign = "middle") %>%
  kable_paper() %>%
  kable_styling(latex_options= c("hold_position")) %>%
  save_kable(here::here("figures/tables/sensitivity_table.jpeg"))
```

# Amphibian Trends

There were 3 amphibian studies comprised of 11 time series in the bioTIME database with sufficient trait coverage to be included in the analysis. With such limited data, we excluded amphibians from the main results, but present preliminary trends here. Models fit with amphibian, mammal, and bird data combined showed qualitatively similar trend results to those reported in the manuscript that did not include amphibians. For models fit with only amphibian data, no general trends were detected for any metric. Table 2 gives model estimates for species-based and species corrected metrics.

```{r resultsTab, echo=FALSE, message = FALSE, warning = FALSE}
readr::read_csv(here::here("figures/amph_model_table.csv")) %>%
  select(metric, everything(), -REML, -bobyqa) %>%
  select(-c(n_ts, formula, model)) %>%
  filter(group != "Residual" | is.na(group), metric %in% c("S", "Jaccard_base", "SES_FRic", "SES_FEve", "SES_FDiv")) %>%
  mutate(metric = ifelse(metric == "Jaccard_base", "log(Jaccard + 1)", metric),
         metric = ifelse(metric == "S", "log(Species Richness)", metric),
         effect = ifelse(effect == "ran_pars", "random", effect),
         group = case_when(
           group == "rarefyID:study_id" ~ "time series within study",
           group == "study_id" ~ "study",
           group == "Residual" ~ "residual",
           TRUE ~ group),
         term = case_when(
           term == "(Intercept)" ~ "Intercept",
           term == "year_scaled" ~ "Year",
           term == "sd__(Intercept)" ~ "SD Intercept",
           term == "cor__(Intercept).year_scaled" ~ "Corr(Intercept, Year)",
           term == "sd__year_scaled" ~ "SD Year",
           term == "sd__Observation" ~ "SD Observation",
           TRUE ~ term
         )) %>%
  mutate(across(c(estimate, std.error, p.value), round, 2),
         p.value = ifelse(p.value == 0, "<0.001", p.value)) %>%
  arrange(metric != "Species Richness", metric, !is.na(group), group, term == "Corr(Intercept, Year)", term) %>%
  dplyr::rename(grouping = group) %>%
  #kable(format = "latex", caption = "Model estimates and statistics for general trend models for species richness, Jaccard similarity, and standardized functional diversity metrics for Amphibian time series.") %>%
  kable(format = "latex") %>%
  column_spec(3, width = "2cm") %>%
  collapse_rows(columns = 1:3, valign = "middle") %>%
  kable_paper() %>%
  kable_styling(latex_options= c("scale_down", "hold_position")) %>%
  save_kable(here::here("figures/tables/amph_table.jpeg"))
```

# Missing Individual Study Models

We fit models separately for each study and metric to estimate study-level trends. 11 models could not be fit because there were to many time series relative to the number of observations, and therefore not enough data to fit the appropriate mixed effects model structure. The missing models are in Table 3.

```{r missingTab, echo=FALSE, message = FALSE, warning = FALSE}
readr::read_csv(here::here("paper/missing_indv_studies.csv")) %>%
  rename(`n time series` = nrarefyID, `n observations` = n) %>%
  #kable(format = "latex", caption = "Model estimates and statistics for general trend models for species richness, Jaccard similarity, and standardized functional diversity metrics for Amphibian time series.") %>%
  kable(format = "latex") %>%
  kable_paper() %>%
  kable_styling(latex_options= c("scale_down", "hold_position")) %>%
  save_kable(here::here("figures/tables/missing_models_table.jpeg"))
```
